MSSM-Penn-Pitt - ACC and DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model

### 2020-01-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
update metadata schema, automate outlier computation

#Get Metadata
metadata = 'syn16816488' 
ALL_USED_IDs[length(ALL_USED_IDs)+1] = metadata
metadata = downloadFile_version(metadata, version = 13)  %>% 
  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch','rnaSeq_dissection:Brain_Region',
                                      'rnaSeq_report:Exclude?',
                                      'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency',
                                      'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned'))

METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                Tissue = `rnaSeq_dissection:Brain_Region`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, 
                IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, 
                GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, 
                rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, 
                AlignmentRate = `rnaSeq_report:Percent_Aligned`,
                RIN = `rnaSeq_isolation:RIN`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
                RIN, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, Tissue, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>% 
  dplyr::filter(Institution %in% c("MSSM", "Penn", "Pitt")) %>% 
  dplyr::filter(SampleID %in% colnames(counts), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))
```

@@

##XIST and UTY expression 
#ENSG00000229807.10 and ENSG00000183878.15 
FILT <- REPORTED.GENDER.COUNTS %>% 
  filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
  dplyr::select(-one_of("chromosome_name")) %>% 
  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
  mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
  mutate(UTY = as.numeric(`ENSG00000183878.15`))

p = ggplot(FILT, aes (x= XIST, y = UTY)) 
p = p + geom_point(aes(color=`Reported Gender`)) 
p
```
### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

@@

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

@@

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```

>>>>

#Get Metadata
metadata = 'syn16816488' 
ALL_USED_IDs[length(ALL_USED_IDs)+1] = metadata
metadata = downloadFile_version(metadata, version = 14)  %>% 
  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch','rnaSeq_dissection:Brain_Region',
                                      'rnaSeq_report:Exclude?',
                                      'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency',
                                      'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned'))

METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                Tissue = `rnaSeq_dissection:Brain_Region`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, 
                IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, 
                GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, 
                rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, 
                AlignmentRate = `rnaSeq_report:Percent_Aligned`,
                RIN = `rnaSeq_isolation:RIN`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                RIN, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, Tissue, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>% 
  dplyr::filter(Institution %in% c("MSSM", "Penn", "Pitt")) %>% 
  dplyr::filter(SampleID %in% colnames(counts), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))
```

@@

##XIST and UTY expression 
#ENSG00000229807.11 and ENSG00000183878.15 
FILT <- REPORTED.GENDER.COUNTS %>% 
  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
  dplyr::select(-one_of("chromosome_name")) %>% 
  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
  mutate(UTY = as.numeric(`ENSG00000183878.15`))

p = ggplot(FILT, aes (x= XIST, y = UTY)) 
p = p + geom_point(aes(color=`Reported Gender`)) 
p
```
### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

@@

# Percentage from each PC
eigen <- PC$sdev^2
pc1 <- eigen[1]/sum(eigen)
pc2 <- eigen[2]/sum(eigen)

# Identify outliers - samples 4SDs from the mean
outliers <- as.character(plotdata$SampleID[c(which(plotdata$PC1 < mean(plotdata$PC1) - 4*sd(plotdata$PC1)),
                              which(plotdata$PC1 > mean(plotdata$PC1) + 4*sd(plotdata$PC1))), drop = T])

outliers <- c(outliers, as.character(plotdata$SampleID[c(which(plotdata$PC2 < mean(plotdata$PC2) - 4*sd(plotdata$PC2)),
                                           which(plotdata$PC2 > mean(plotdata$PC2) + 4*sd(plotdata$PC2))), drop = T] ))

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
    tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID) %>% 
  dplyr::mutate(label = ifelse((label %in% outliers), label, NA))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

@@

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% outliers),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), outliers)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

@@

Based on the expression pattern following samples were tagged as outliers: `r paste(outliers, collapse = ', ')`

Based on prior evidence, investigators elected to remove: `r setdiff(indToRemove, outliers)[setdiff(indToRemove,outliers) %in% colnames(counts)]`

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'ageOfDeath', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```
